Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-10-20 text/ms.tex
Arrange files into analysis and text folders
Develop introduction

Black swans are unexpected extreme events with dramatic consequences. One of
the most striking black swans in ecology was the asteroid marking the mass
extinction of the K-T boundary (REF). Today, it is the extremeness of climate
--- particularly in concert with shifts in mean temperature --- that is
expected to cause the greatest societal damage (REF).

However, while extremes in the physical environment are present and widely
accepted, it remains unclear the extent to which ecological systems buffer or
suffer from black swans.

Existing evidence for black swans or heavy tails in population dynamics is thus
far limited. Considerable work has focussed on heavy tails in physical
environmental processes in, for example, rain fall (REF), wave height (REF),
earth quake magnitude, climate temperature (REF). time to extinction, plankton
time series, GPDD work found little evidence subsequent work suggested, but...
Ward applied a data-intense method and found evidence for.

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not exist. Indeed, the majority of population
dynamics model fitting and risk forecasting assumes that population dynamics
are not heavy tailed (e.g.~REF, REF, REF). Alternatively, population dynamics
might be heavy tailed, but a paucity of statistical methods and shortness of
available time series has limited our ability to detect them. For example,
a commonly applied probability distribution to detect the shape of time series
tails (the generalized extreme value distribution) requires lumping data into
the most extreme event per unit time (e.g.~heaviest rainfall per year) and
therefore requires a considerable quantity of data (REF). Previous work into
the shape of population dynamic tails, has focused on model comparison to
identify parsimonious explanations from a set of particular hypotheses, but
does not quantify the magnitude of heavy tails on a continuous scale or put the
conclusions in a probabilistic framework (REF). Furthermore, information
criteria are poor at distinguishing subtle differences in population dynamics
for the relatively short time series that are most readily available (REF CHANGE
THIS?).

Or bring back in Ward - data intensive, can't be applied to most individual
time series

plankton --- hideous distribution to fit, even harder in population dynamics
framework -and what about seasonal cycles, density dependence?

GPDD - model competition and finding most parsimonious explanation from set of
hypotheses --- but only \emph{very} heavy from not heavy

GEV --- super data intensive

tipping point and regime shift methods --- say something\ldots{} not
appropriate for most of the data available

@@

Papers to work in: \citep{inchausti2002, halley2002, inchausti2001}
\citep{jentsch2007} \citep{ward2007} \citep{garcia-carreras2011}
\citep{sornette2009} \citep{nunez2012} \citep{thompson2013}
\citep{beaugrand2012} \citep{pine-iii2009} \citep{doak2008} Extreme rain and
monsoons over India increasing \citep{goswami2006} \citep{smale2013}
\citep{easterling2000} \citep{scheffer2003} \citep{katz2005} \citep{taleb2007}
\citep{vasseur2014} \citep{vert-pre2013} \citep{lindenmayer2010}
\citep{valpine2002} \citep{gregory2010} \citet{garcia-carreras2011}
\citeauthor{brook2006} \citetext{\citeyear{brook2006};
  \citealp{herrandoprez2014}} \citep{sibly2005, ziebarth2010}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{t-nu-eg.pdf}
\caption{
An illustration of fitting population dynamics models that allow for heavy
tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
The probability density for t distributions with a scale parameter of 1 and
different values of $\nu$. Small values of $\nu$ create heavy tails. As $\nu$
approaches infinity the distribution approaches the normal distribution. For
example, at $\nu = 2$, the probability of drawing a value less than -5 is
1.8\%, whereas the probability of drawing such a value from a normal
distribution is nearly zero ($2.9\cdot10^{-5}$\%). (c--e) Simulated population
dynamics from a Gompertz model with process noise drawn from t distributions
with different values of $\nu$. Coloured dots in panels c and d represent jumps
with less than a 1 in 1000 chance of occurring in a normal distribution. (f--h)
Estimates of $\nu$ from models fit to the times series in panels c--e. Shown
are the posterior samples (histograms), median and interquartile range of the
posterior (IQR) (dots and line segments), and the exponential prior on $\nu$
(dashed lines). Colour shading behind panels f--h illustrates the region of
heavy tails.}
\label{fig:didactic}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.44\textwidth]{nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\
populations of birds, mammals, insects, and fishes. Small values of $\nu$
($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest
approximately normal-tailed dynamics. Vertical points and line segments
represent posterior medians and 50\% / 90\% credible intervals for individual
populations. Inset plots show probability that $\nu < 10$ (probability of heavy
tails) for populations arranged by taxonomic order. Taxonomic orders are sorted
by decreasing mean Pr($\nu < 10$). Taxonomic orders with three or fewer
populations in panel a are omitted for space. Red and orange colours highlight
populations with a high probability of heavy-tailed dynamics.
}
\label{fig:nu-coefs}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{correlates-p10.pdf}
\caption{
Potential correlates of heavy-tailed behaviour (indicated by a high probability
that $\nu < 10$). Shown are (a--c) parameters from the Gompertz heavy-tailed
population model, (b) number of time steps, (c) body length, and (d) lifespan.
For the Gompertz parameters, $\sigma$ refers to the scale parameter of the
Student-t process-noise distribution, $\lambda$ refers to the population growth
rate at an abundance of one, $b$ refers to the density dependence parameter
($1$ is maximally density independent, $0$ is maximally density dependent, and
$<0$ is inversely density dependent).
%For parameter values, points and lines represent medians and 50\% credible
%intervals of the posterior.
Circles representing a limited number of sharks, crustaceans, and gastropods
are filled in white. Median and 90\% credible interval posterior predictions of
a beta regression multilevel model are shown in panels a and d where there was
a high probability the slope coefficient was different from zero (Supporting
Material).
}
\label{fig:correlates}
\end{center}
\end{figure}

>>>>

Black swans are unexpected extreme events with dramatic consequences
\citep{taleb2007,sornette2009}. One of the most striking black swans in ecology
was the asteroid marking the mass extinction of the K-T boundary (REF). Today,
it is the extremeness of climate, in concert with shifts in mean temperature,
that is expected to cause the greatest societal damage \citep{ipcc2012}.
However, whereas extremes in the physical environment such as wave height, rain
fall, and temperature are present and widely accepted
\citep{gaines1993,katz2005}, it remains unclear the extent to which ecological
systems buffer or suffer from black swans \citep{nunez2012}.

Existing evidence for black swans or heavy tails in population dynamics is
mostly anecdotal or for particular systems. For example, \citet{doak2008}
reviewed a collection of field experiment population dynamic surprises and
conclude that these are more common than we typically assume. \citet{nunez2012}
compiled a list of major events on geologic and contemporary scales that could
be considered ecological black swans. Other work has identified heavy-tail
dynamics for particular systems. For example, \citet{segura2013} identified
heavy tails in time series of marine microbe populations and \citet{drake2014}
identified heavy tails in time to extinction for experimental waterflea
populations. However, systematic evidence has remained elusive.
\citet{keitt1998} tested for power-law tail distributions in breeding bird
abundance, but later worked showed the apparent heavy tails were a result of
mixing populations of different species \citep{allen2001}. A large scale
analysis of the Global Population Dynamics Database (GPDD) by
\citet{halley2002} found little evidence for heavy-tail distributions in longer
abundance time series.

%Considerable work has focussed on heavy tails in physical environmental
%processes in, for example, rain fall (REF), wave height (REF), earth quake
%magnitude, climate temperature (REF). time to extinction, plankton time series,
%GPDD work found little evidence subsequent work suggested, but... Ward applied
%a data-intense method and found evidence for.

%Ecological population dynamic models typically assume 'thin tailed'
%distributions, such as the normal distribution on a multiplicative scale (e.g.
%REF). We define black swan dynamics as cases where jumps in population
%abundance from time step to time step. from heavy-tailed distributions.

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not exist. Indeed, the majority of population dynamics
model fitting and risk forecasting assumes that population dynamics are not
heavy tailed (e.g.~REF, REF, REF). Alternatively, dynamics might be heavy
tailed, but a paucity of statistical methods for the frequently available short
time series has limited our ability to detect them. For example, a commonly
applied probability distribution to detect the shape of time series tails (the
generalized extreme value distribution) requires lumping data into the most
extreme event per unit time (e.g.~heaviest rainfall per year) and therefore
requires a considerable quantity of data (REF). Previous large-scale work on
the distribution of abundance trends focused on model comparison to identify
parsimonious descriptions, but has not quantified the magnitude of heavy tails
on a continuous scale (REF). Furthermore, information criteria are poor at
distinguishing subtle differences in population dynamics for the relatively
short time series that are most readily available (REF CHANGE THIS?).
%bring in Ward - great method but data intensive, can't be applied to most
%individual time series from the GPDD
%plankton --- hideous distribution to fit, even harder in population dynamics
%framework -and what about seasonal cycles, density dependence?
%GPDD - model competition and finding most parsimonious explanation from set of
%hypotheses --- but only \emph{very} heavy from not heavy
%GEV --- super data intensive
%tipping point and regime shift methods --- say something\ldots{} not
%appropriate for most of the data available

@@

% Papers to work in: \citep{inchausti2002, halley2002, inchausti2001}
% \citep{jentsch2007} \citep{ward2007} \citep{garcia-carreras2011}
% \citep{sornette2009} \citep{nunez2012} \citep{thompson2013}
% \citep{beaugrand2012} \citep{pine-iii2009} \citep{doak2008} Extreme rain and
% monsoons over India increasing \citep{goswami2006} \citep{smale2013}
% \citep{easterling2000} \citep{scheffer2003} \citep{katz2005} \citep{taleb2007}
% \citep{vasseur2014} \citep{vert-pre2013} \citep{lindenmayer2010}
% \citep{valpine2002} \citep{gregory2010} \citet{garcia-carreras2011}
% \citeauthor{brook2006} \citetext{\citeyear{brook2006};
%   \citealp{herrandoprez2014}} \citep{sibly2005, ziebarth2010}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/t-nu-eg.pdf}
\caption{
An illustration of fitting population dynamics models that allow for heavy
tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
The probability density for t distributions with a scale parameter of 1 and
different values of $\nu$. Small values of $\nu$ create heavy tails. As $\nu$
approaches infinity the distribution approaches the normal distribution. For
example, at $\nu = 2$, the probability of drawing a value less than -5 is
1.8\%, whereas the probability of drawing such a value from a normal
distribution is nearly zero ($2.9\cdot10^{-5}$\%). (c--e) Simulated population
dynamics from a Gompertz model with process noise drawn from t distributions
with different values of $\nu$. Coloured dots in panels c and d represent jumps
with less than a 1 in 1000 chance of occurring in a normal distribution. (f--h)
Estimates of $\nu$ from models fit to the times series in panels c--e. Shown
are the posterior samples (histograms), median and interquartile range of the
posterior (IQR) (dots and line segments), and the exponential prior on $\nu$
(dashed lines). Colour shading behind panels f--h illustrates the region of
heavy tails.}
\label{fig:didactic}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.44\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\
populations of birds, mammals, insects, and fishes. Small values of $\nu$
($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest
approximately normal-tailed dynamics. Vertical points and line segments
represent posterior medians and 50\% / 90\% credible intervals for individual
populations. Inset plots show probability that $\nu < 10$ (probability of heavy
tails) for populations arranged by taxonomic order. Taxonomic orders are sorted
by decreasing mean Pr($\nu < 10$). Taxonomic orders with three or fewer
populations in panel a are omitted for space. Red and orange colours highlight
populations with a high probability of heavy-tailed dynamics.
}
\label{fig:nu-coefs}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{../analysis/correlates-p10.pdf}
\caption{
Potential correlates of heavy-tailed behaviour (indicated by a high probability
that $\nu < 10$). Shown are (a--c) parameters from the Gompertz heavy-tailed
population model, (b) number of time steps, (c) body length, and (d) lifespan.
For the Gompertz parameters, $\sigma$ refers to the scale parameter of the
Student-t process-noise distribution, $\lambda$ refers to the population growth
rate at an abundance of one, $b$ refers to the density dependence parameter
($1$ is maximally density independent, $0$ is maximally density dependent, and
$<0$ is inversely density dependent).
%For parameter values, points and lines represent medians and 50\% credible
%intervals of the posterior.
Circles representing a limited number of sharks, crustaceans, and gastropods
are filled in white. Median and 90\% credible interval posterior predictions of
a beta regression multilevel model are shown in panels a and d where there was
a high probability the slope coefficient was different from zero (Supporting
Material).
}
\label{fig:correlates}
\end{center}
\end{figure}
